---
title: "SummarizingData_filled"
author: "Amy King-Henry"
date modified: "September 2, 2026"
format: html
editor: visual
---

# Miscellaneous Data Summarizing Skills

### Components of a boxplot

- The middle line is the **median**.

- The box is the **interquartile range, or IQR** - the bounds of the second and third quartiles.

- The upper whisker extends from the hinge to the **largest value no further than 1.5 \* IQR** from the hinge (where IQR is the inter-quartile range, or distance between the first and third quartiles).

- The lower whisker extends from the hinge to the **smallest value at most 1.5 \* IQR** of the hinge.

- Data beyond the end of the whiskers are called "outlying" points and are plotted individually.

These are NOT mean and standard deviation!

### Summary Statistics for Continuous Variables

```{r}
# This chunk will not run
mean() : Mean
min() : minimum value
max() : maximum value
range() : min and max values
var() : Variance
sd() : Standard deviation
sd()/mean() : Coefficient of Variation
sd()/sqrt(n()) : Standard error
n() : sample size/count of rows
median() : Median
IQR() : Interquartile Range
quantile() : quantiles
summary() : min, 1st quartile, median, mean, 3rd quartile, max
```

```{r}
load(file = "Demo Data Files/CCdata_tidy.Rdata")
install.packages("car")
library(tidyverse)
```

### Example using summarize to make a nice summary table of some old owl limpet data:

```{r}
Size_summary <-  CCdata_tidy |>  
  group_by(Impact, Year, Season, Survey_Date) |>  
  summarize(
    mean_size = mean(Size, na.rm = TRUE), # mean 
            sd_size = sd(Size, na.rm = TRUE),  # standard deviation
            n = n(), # count
            se_size =  sd_size/sqrt(n)) |> # standard error
  mutate_at(vars(mean_size, sd_size, se_size), # bonus line to reduce digits
            .funs = round, 2) 

Size_summary |> 
  select(-Survey_Date) |> 
  knitr::kable()
```

#### Example creating a plot from that summary data:

```{r}
Size_summary |> 
  ggplot() + 
  aes(x = as.factor(Survey_Date), y = mean_size, fill = Impact) + 
#  geom_point(position = position_dodge(0.9)) +
geom_col(position = position_dodge()) +
  geom_errorbar(aes(ymin = mean_size - se_size,
                    ymax = mean_size + se_size), 
                position = position_dodge(0.9),
                width = 0.2)
```

#### Beyond the summary stats: frequency distributions

Boiling it down to the mean and standard deviation really loses a lot of information though. It's always good to look at the histogram/distribution to get a better sense of what's going on. Let's look back at just that one histogram.

```{r}
CC_sp19 <- CCdata_tidy |> 
  filter(Season == "Spring" & Year == 2019 & Impact == "High")

CC_sp19 |> 
  ggplot() +
  aes(x = Size) +
  geom_histogram(aes(y = after_stat(density))) 
  
CC_sp19 |> 
  ggplot() +
  aes(x = Size) +
  geom_density()
```

What observations can we make about the distribution of this data?

- Is it symmetrical or skewed?

- Are there many outliers?

- Is there missing data?

- Is it unimodal, i.e., is there only one peak where there are the most observations, or are there multiple peaks (bimodal, etc.)?

- What are our expectations for what size demographics should look like for populations of:

  - long-lived vs. short-lived animals

  - r vs. K-selected

  - growing vs. shrinking populations?

- What factors apart from the population itself may have influenced this pattern? Let's look at the histograms for all the dates and sites:

```{r many histograms}
CCdata_tidy |> 
  ggplot() +
  aes(x = Size) +
  geom_histogram() +
  facet_wrap(Survey_Date ~ Impact)
   


```

On the whole, how would you describe these distributions?

#### The Normal Distribution

Many statistical tests and linear models are based on assumptions that data is normally distributed! Therefore, one thing we'll do a lot of is assess distributions and decide if they can be reasonably approximated by the normal distribution or not.

This applies only to data within a sample, but in the context of a statistical test we will apply tests of normality to the residuals, or the differences from the observed and expected values.

How do we assess?

1.  Visually. Is it symmetrical? Are most of the observations in the middle?

2.  With a QQ plot. Do the observations mostly fall close to the 1:1 line?

3.  With a Shapiro-Wilk test (we'll learn this later)

Visual:

```{r}
CC_sp19 <-  CCdata_tidy |> 
  filter(Season == "Spring" & Year == 2019 & Impact == "High")

CC_sp19 |> 
  ggplot() +
  aes(x = Size) +
  geom_histogram(aes(y = after_stat(density)))
```

2.  With a QQ plot

```{r}
# here's one function from Base R
qqnorm(CC_sp19$Size)

# And another from the car package that has 95% confidence intervals
car::qqPlot(CC_sp19$Size)
```

Next week, we will start learning about statistical tests which can quantify and assign a probability to how closely your data fits a particular type of distribution.

If they are normal, then we can use a class of tests called parametric tests. If not, we use non-parametric tests.

## Assessing a MODEL for normality

It's one thing to assess a single sample for normality, but when we do a statistical test, we don't actually worry too much about the data in the samples themselves - what matters is the normality of the residuals, or the individual differences of Observed - Expected.

Here is an example dataset of masses of organisms taken from two populations.

```{r}
mass_g1 <- exp()
mass_g2 <- e
mass_g <- c()

fakedata <- tibble(Group = )

```

We ultimately plan to do a t-test, but the easiest way to assess for normality is to make a simple linear model and make some automatic residual plots.

```{r}
model <- lm()
plot()

```

The data itself also does look skewed, but it's important that we view it separately for the two samples:

```{r}
ggplot() + 
  aes(x = , y = ) + 


# the variance is really skewed! 
```

```{r}
ggplot() +  
  facet_wrap()
```

Maybe a log transformation will help!

To log transform this data, we can create a new column or just apply the transformation in the ggplot or model.

```{r}
fakedata <-  |> 
  mutate()

log_model <- lm()

plot()

```

Those diagnostic plots look much better!

The data look more like normal histograms when you plot the samples separately as well.

```{r}
ggplot() + 
  aes() + 
  geom_histogram() + 
  facet_wrap()
```
